---
title: "Statistics with R"
subtitle: "Data Wrangling"
author: "Zhuanghua Shi (Strongway)"
date: "14 May 2018"
output: 
  beamer_presentation: 
    includes: 
      in_header: preamble.tex
citation_package: natbib
link-citations: yes
latex_engine: xelatex
---

```{r setup, include=FALSE}
library(knitr)
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)
library(tidyverse)
```

# Dealing with data

After you collect your raw data, you often need to:

1. tidy data (remove/replace missing data)
2. define outliers and refine your data
3. select subset of your data
4. join different data together
5. summarize your data
6. compute new type of data

# Manipulate data with Tidyverse

* Unlike traditional approach, `tidyverse` use pipes, which is close to our logical processing. 

> Pipe operation: %>%

  Data %>% Operator() %>% Operator() %>% ... -> Results

# Most used operators
  * `filter()` - pick rows based on conditions
  * `group_by()` - group rows of observations together
  * `summarize()` - compute summary measures, such as mean, sd, count etc.
  * `mutate()` - create new variable (column)
  * `arrange()` - sort the data based on a variable
  

> data %>% group_by(condition) %>% summarise(mrt = mean(RT))

# The filter verb

`filter()` select some rows 

\fig{0.7}{filter.pdf}

# Example

```{r, echo = T}
library(gapminder)
gapminder %>% filter(year == 1952, country == 'Germany')
```

# The arrange verb

`arrange()` sorts a table based on a variable

\fig{0.7}{sort.pdf}

# Example: Sorting with arrange

```{r, echo = T}

gapminder %>% arrange(gdpPercap)

```

# Example: Sorting with arrange

* decending using `desc()`

```{r, echo = T}

gapminder %>% arrange(desc(gdpPercap))

```

# The mutate verb

`mutate()` changes or adds variables

\fig{0.7}{mutate.pdf}

# Using mutate to change a variable

```{r, echo = T}
gapminder %>% mutate(pop = pop/1000000)
```

# Using mutate to add a new variable

```{r, echo = T}
gapminder %>% mutate(gdp = gdpPercap * pop)
```

# The summarize verb

```{r, echo = T}
gapminder %>% 
  summarize(meanlifeExp = mean(lifeExp))

```

* Functions you can use for summarizing
    * mean / median
    * sum
    * sd
    * min/max

# The group_by verb

* group_by verb is useful if you want to summarize different groups

```{r, echo = T}
gapminder %>% 
  group_by(year)%>%
  summarize(meanlifeExp = mean(lifeExp))
```

# Join tables

Please check the data manipulation cheat sheet. 

* left_join(x,y, by = )
* right_join(x,y, by = )
* inner_join(x,y, by=)
* full_join(x,y, by = )

# Data manipulation example

In the following practice, we will do a full process on data manipulation. The raw data are availabe in our shared github. 



