---
title: "Statistics with R"
subtitle: "Linear Regression"
author: "Zhuanghua Shi (Strongway)"
date: "11 June 2018"
output: 
  beamer_presentation: 
  includes: 
  in_header: preamble.tex
citation_package: natbib
link-citations: yes
latex_engine: xelatex
---

```{r setup, include=FALSE}
library(knitr)
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
library(tidyverse)
library(broom)
library(janitor)
library(ez)
```


# Linear regression

* A continuous dependent variable $Y$
* One of more independent variables $X$

We want to estimate their linear relationship:

$$ Y = b_0 + b_1 X + \epsilon$$

# An example

* build-in `cars` dataset

```{r}
ggplot(cars, aes(speed, dist)) + geom_point() + 
  geom_smooth(method = 'lm')
```


# Build linear model

* Linear regression uses function `lm()`
* `lm()` accept
    * formula
    * data

```{r}
mod1 = lm(dist ~ speed, data = cars)
mod1
```

# Get model summary

* summary(model) provides residuals, coefficients, statistics, significances, R-squared, F-tests

```{r}
summary(mod1)
```

# Goodness of fit

* How do we know if the model is good fit?
    * R-Square: higher the better (>0.7)
    * t-statistics: p-value should be less than 0.05

# Prediction of the model

* `predict(model, testData)`
    * note: testData shold contain the IV variables

```{r}
p_dist = predict(mod1, data.frame(speed = c(20,21)))
print(p_dist)
```

# R formula

Recall the last session:

$$Y \sim X1 + X2 + 1$$
$$ Y \sim X1*X2 $$

$$ Y \sim X1 - 1$$

# Example of model comparison

* Data of Thibault et al. (2007) motion sensitity as a function of age, movement type (First/Second order motion), Sex

```{r}
motion = read.csv('motion.csv')
head(motion,3)
```

# Explore the data 
* Explore the relation between the age and motion threshold
    * Separate for motion type and gender

```{r}
motion %>% ggplot(aes(LnAge, LnThresh)) + geom_point() + 
  facet_grid(Mtype~Sex)
```

# Build linear models

* Only main factors
```{r}
motion_mod1 = lm(LnThresh ~ Mtype + Sex + LnAge, data = motion)
coef(summary(motion_mod1))
```

# Build linear models

* Main factors and two-way interaction

```{r}
motion_mod2 = lm(LnThresh ~ Mtype + Sex + LnAge + Mtype:Sex +
                   Mtype:LnAge + Sex:LnAge, data = motion)
coef(summary(motion_mod2))
```


# Compare two models

* Using ANOVA method to compare the two models, with and without the second-order interaction

```{r}
anova(motion_mod1, motion_mod2)
```

# Remove insig. factor

* `Sex` is not a critical factor, shown by the t-test

```{r}
motion_mod3 = lm(LnThresh ~ Mtype +  LnAge, data = motion)
coef(summary(motion_mod3))
```

# Compare models

* Compared with and without the factor `Sex`
    * Removal of `Sex` did not make any difference

```{r}
anova(motion_mod1, motion_mod3)
```

# Visualize the best model

* `lm` method provide direct `plot` for diagnose

```{r}
plot(motion_mod3)
```

# Linear model and tidyverse

* Example - Modelfest data from Watson & Ahumada (2005)
    * Foveal detection of spatial contrast
    * http://jov.arvojournals.org/5/9/6
    
* Using RStudio 'File - Import Dataset' to import

```{r}
library(readxl)
url <- "http://jov.arvojournals.org/data/Journals/JOV/932836/modelfestbaselinedata.xls"
destfile <- "modelfestbaselinedata.xls"
curl::curl_download(url, destfile)
ModelFest <- read_excel(destfile, col_names = FALSE) %>% 
  clean_names() # from janitor package, make names accessible
```
# Tidy the data

```{r}
library(tidyr)
dat = gather(ModelFest,, threshold, -x_1) 
dat$stim = rep(1:43, each = 4*16) # 43 stimuli, 4 repetition, 16 subjects
# first 10 stimuli spatial frequency
SpatFreq <- c(1.12, 2^seq(1, 4.5, 0.5), 30)

# only analysis for the first 10 stimuli
dat %>% filter(stim <=10) %>% mutate(freq = SpatFreq[stim]) -> dat1

ggplot(dat1, aes(freq, threshold, color = x_1, group = x_1)) + 
  geom_point() + geom_line()
```

# some useful tricks get model parameters

* code from moderndive.com

```{r, eval = FALSE}
library(broom)
library(janitor)
model %>% tidy(conf.int = TRUE) %>%
  mutate_if(is.numeric, round, digits = 3) %>%
  clean_names()
```

* `glance()` 


# Model with multiple subsets

```{r}
lmod <- function(df){
  lmfit = lm(threshold ~ freq, data = df)
  return(lmfit)
}

dat1 %>% group_by(x_1) %>% nest() %>%
  mutate(mod = map(data, lmod)) %>% 
  mutate(glance = map(mod, broom::glance)) %>%
  unnest(glance) -> gm

ggplot(gm, aes(x_1, r.squared)) + geom_point()  
```

# References

Some contents from this session are from:

1. Knoblauch & Maloney, Modeling Psychophysical Data in R, 2012
2. An introduction to statiscal and Data Science via R, Moderndive.com
