---
title: "Statistics with R"
subtitle: "Logistic Regression"
author: "Zhuanghua Shi (Strongway)"
date: "18 June 2018"
output: 
  beamer_presentation: 
  includes: 
  in_header: preamble.tex
citation_package: natbib
link-citations: yes
latex_engine: xelatex
---
  
```{r setup, include=FALSE}
library(knitr)
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
library(tidyverse)
library(broom)
library(janitor)
library(ez)
```


# Logisitic regression

* One of more independent variables $X$
* A binary output $Y$

Instead of model $Y$ directly, we model the log odds of the event 

$$Z = \ln\frac{P}{1-P}$$
where $P$ is the probability of event. We want to estimate the relationship:

$$ Z = b_0 + b_1 X + \epsilon$$

The above equation can be modeled using the `glm()` function by setting `family` to `binomial`. 

# An example

* Ternus apparent motion
    * typical two types of motion percepts - Group vs. Element motion
* The motion percept is mainly determined by the inter-stimulus interval (ISI)
    * https://en.wikipedia.org/wiki/Ternus_illusion
* Chen et al. conducted one experiment on how auditory trains influence on visual apparent motion
    * source available at github: https://github.com/msenselab/temporal_averaging
    

# Visualize the data

First, let's visualize the data

```{r fig.height=2, fig.width=3}
dat = read.csv('https://raw.githubusercontent.com/msenselab/temporal_averaging/master/exp1.csv')

dat %>% group_by(mIntv, soa) %>%
  summarise(p = mean(resp)) %>%
  ggplot(aes(soa, p, color = mIntv, group = mIntv)) +
    geom_point() + geom_line()
```



# Build logistic model

* Logistic regression uses function `glm()`
    * `family` objects provide a convenenient way to specify the details of the models: binomial, gaussian, Gamma, inverse.gaussian, poisson, quasibinomial et. 
    * Within the `family`, you can link a specific function, e.g.,
    
> `binomial(link = 'logit')
> `binomial(link = 'probit')

```{r}
dat %>% filter(sub == 1, mIntv ==1) %>% 
  group_by(soa) %>% summarise(p = mean(resp)) ->sub11
gsub1= glm(p ~ soa, family = binomial(link = 'logit'), data = sub11)
kable(tidy(gsub1))
```
# fit from raw data

* We can fit the logistic model directly from the raw data
    * What is the difference output this time? why?

```{r}
dat %>% filter(sub == 1, mIntv ==1) -> sub1
gsub2 = glm(resp ~ soa, data = sub1, family = binomial(link = 'logit'))
kable(tidy(gsub2))
```

# Interpreting the parameters

$$\ln\frac{P}{1-P} = b_0 + b_1 \cdot X$$

* When $P = 0.5$, we have $X_{0.5} = -b_0/b_1$ as the threshold (PSE). 

* When $P = 0.75$, $X_{0.75} = (ln3 - b_0)/b_1$. So we can obtain the JND: $JND = ln3/b_1$

# Build a function for tidyverse

* Using `glm()` build a function
    * doing logistic regression
    * estimate parameters
    * calculate PSE and JND
    
```{r}
fitPsy <- function(df){
  glm(resp ~ soa, family = binomial(link = 'logit'), data = df) %>%
    tidy(.) %>% select(term, estimate) %>%
    spread(term, estimate) %>% 
    rename(b1 = soa, b0 = `(Intercept)`) %>% 
    mutate(pse = -b0/b1, jnd = log(3)/b1)
}

```

# Build a pipe for analysis

* nest each subject, each condition
* estimate logistic function
* return the paramters

```{r}
dat %>% group_by(sub, mIntv) %>%
  nest() %>% 
  mutate(est = map(data, fitPsy)) %>%
  unnest(est, .drop = TRUE) -> thresholds
```

# Repeated-measures ANOVA

* Test if the auditory train has any impact on visual motion percept.
```{r}
av1 = ezANOVA(data = thresholds,
        dv = pse,
        wid = sub, 
        within = mIntv)
kable(av1$ANOVA)
```


# Visualize the thresholds

```{r fig.height=2, fig.width=3}
thresholds %>% group_by(mIntv) %>%
  summarise(mpse = mean(pse), n = n(),
            se = sd(pse)/sqrt(n-1)) %>%
  ggplot(aes(mIntv, mpse)) + geom_bar(stat = 'identity', width = 0.5) + 
  geom_errorbar(aes(ymin = mpse - se, ymax = mpse+se), width = 0.2) + 
  coord_cartesian(ylim = c(100, 180)) + theme_classic() + 
  xlab('Auditory Intervals') + ylab('Threshold (ms)')
```

# Second approach with `quickpsy`

* `quickpsy` by Daniel Linares
    * http://dlinares.org/quickpsy.html
    
A package for quickly fitting and plotting psychometric functions. 

* Fits and plots multiple conditions
* Calculates parametric and non-parametric bootstrap confidence intervals
* Guess and lapses can be fixed or free as parameters
* Performs goodness-of-fit
* compute AIC

# QuickPsy

* QuickPsy uses a general psychometric function

$$\Psi(x) = \gamma + (1-\gamma - \lambda) f(x)$$
where $\gamma$ is the guess rate, $\lambda$ is the lapse rate, and $f(.)$ is a sigmoidal-shape function. 

```{r fig.height=2, fig.width=3}
library(quickpsy)

pest = quickpsy(sub1, x = soa, k = resp)
plotcurves(pest)
```

# Paramters of quickpsy

* grouping: name of the grouping variables, accept multiple conditions 
* within: name of the within-factor variable
* between: name of the between variable
* fun: name of the shape of the curve. (cum_normal_fun, logistic_fun, weibull_fun)

```{r, eval=FALSE}
fit <- quickpsy(Vernier, Phaseshift, NumUpward, N,
                grouping = .(Direction, WaveForm, TempFreq))
```

# Using quickpsy with tidyverse

```{r fig.height=3, fig.width=5}
quickpsy(dat, soa, resp, grouping = .(sub), 
         within = .(mIntv), bootstrap = 'none') -> pses
plotcurves(pses)
```

# Visualize the thresholds

* output of the quickpsy has multiple tables
    * .$thresholds contain the estimated thresholds
    
```{r fig.height=2, fig.width=3}
pses$thresholds %>% group_by(mIntv) %>%
  summarise(mpse = mean(thre)) %>%
  ggplot(aes(mIntv, mpse)) + geom_bar(stat = 'identity', width = 0.5)
```

# Let's practice

* Next two weeks - Jan Nasemann will provide related tutorials

